---
title: "Lab 2"
author: Filippo Gambarota
format: html
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      dev = "svg")
```

```{r packages, message=FALSE, warning=FALSE}
devtools::load_all()
library(here)
library(tidyr) # for data manipulation
library(dplyr) # for data manipulation
library(ggplot2) # plotting
library(performance) # diagnostic
library(car) # general utilities
library(MuMIn) # model selection
library(patchwork)
```

```{r}
#| label: ggplot
#| include: false

mytheme <- function(size = 15){
  theme_minimal(size)
}

theme_set(mytheme())
```

# Overview

```{r}
#| echo: false  
data("volunt")
```


The `volunt` dataset contains data from `r nrow(volunt)` people on volunteering.

The dataset contains:

- `id`: identifier for each person
- `age`: age of the participants in years
- `ses`: socioeconomic status, from 1 (low) to 4 (high)
- `sconnection`: a scale measuring the social connection within the community. From 1 (low) to 10 (high)
- `vol`: is the person involved in volunteering? 1 = yes, 0 = no

# Steps

1. Import the data
2. Explore the dataset
    - univariate distributions of the variables
    - bivariate relationships
3. fit a model testing all main effects and interpret the parameters
4. fit a model testing the interaction between ses and social connection. Compare the model with the previous model and interpret the result.
