---
title: "Lab 6"
author: "Filippo Gambarota"
format: html
bibliography: "`r filor::fil()$bib`"
embed-resources: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

devtools::load_all()
library(tidyverse)
library(ggeffects)
```

# Overview

# EDA

## Loading data

```{r}
data("psych")
dat <- psych
```

Given that we did not introduced random-effects models, we select a single subject to analyze.

```{r}
cc <- seq(0, 1, 0.1)

dat$contrast_c <- cut(dat$contrast, cc, include.lowest = TRUE)

dat |> 
  filter(id %in% sample(unique(dat$id), 5)) |> 
  group_by(id, cond, contrast_c) |> 
  summarise(y = mean(y)) |> 
  ggplot(aes(x = contrast_c, y = y, color = cond, group = cond)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~id)
```

```{r}
dat <- filter(dat, id == 6)
```

We have several interesting stuff to estimate. Let's start by fitting a simple model:

```{r}
fit1 <- glm(y ~ contrast, data = dat, family = binomial(link = "logit"))
summary(fit1)
```

The parameters `(Intercept)` and `contrast` are respectively the probability of saying yes for stimuli with 0 contrast. Seems odd but in Psychophysics this is a very interesting information. We can call it the false alarm rate. We usually expect this rate to be low, ideally 0.

```{r}
plogis(coef(fit1)[1])
```

The `contrast` is the slope i.e. the increase in the log odds of saying yes for a unit increase in `contrast`. In this case this parameter is hard to intepret, let's change the scale of the contrast:

```{r}
dat$contrast10 <- dat$contrast * 10
fit1 <- glm(y ~ contrast10, data = dat, family = binomial(link = "logit"))
summary(fit1)
```

Now the `contrast10` is the increase in the log odds of saying yes for an increase of 10% contrast. Using the divide-by-4 rule we obtain an maximal increase of `r coef(fit1)[2]/4` of probability of saying yes.

Another interesting parameter is the threshold. In psychophysics the threshold is the required $x$ level (in this case `contrast`) to obtain a certain proportions of $y$ response.

For a logistic distribution [see @Knoblauch2012-to] the 50% threshold can be estimated as $-\frac{\beta_0}{\beta_1}$ thus:

```{r}
-(coef(fit1)[1]/coef(fit1)[2])
```

Then the slope is simply the inverse of the regression slope and represent the increase in performance/visibility for a unit increase in $x$:

```{r}
1/coef(fit1)[2]
```

In fact, we can use these parameters to plot a logistic distribution:

```{r}
curve(plogis(x, -(coef(fit1)[1]/coef(fit1)[2]), 1/coef(fit1)[2]),
      0, 10)
```

That is very similar to the effects estimated by our model:

```{r}
plot(ggeffect(fit1))
```

# References
